---
title: "Business Analytics II - Group 1 - Home Assignment 3"
author: "Yurii Masiuk, Vanda Lakatos, Kornel Auer, Vivien Arns, Benjamin Meixner, Nicholas Olsa"
date: "2024-11-30"
output: 
  pdf_document: 
    latex_engine: tectonic
    fig_caption: yes
---

## Completed exercises: all (1,2)

```{r, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


```{r}
library(xts)
library(lubridate)
set.seed(69420)
```

# Exercise 1 

First, we call part of the code from class 3 that we will reuse for the exercises. 

```{r}
setwd("/home/maixnor/repo/wu/r/ba2/data")
price_data <- read.csv(file="price_data.csv", sep = ";", dec = ",")
firm_data <- read.csv(file="firm_data.csv", sep = ";", dec = ",")

price_data <- na.omit(price_data)
firm_data <- na.omit(firm_data)

IDs <- unique(price_data$ID)

# Creating xts object for prices of all firms
IDs <- unique(price_data$ID)
head(IDs)
p <- list()
for (i in 1:length(IDs)) { 
  price_data_i <- price_data[price_data$ID==IDs[i], ]
  p[[i]] <- xts(price_data_i$Price,
       order.by=as.Date(as.character(price_data_i$Date), format="%Y-%m-%d"))
}
price <- do.call(cbind, p)
colnames(price) <- paste0("X", IDs)
dim(price)
price <- price["1990/"]
price[1:10,1:5]

price <- apply.daily(price, FUN=function(x) x[1,] )
dim(price)
price <- abs(price)
price <- na.approx(price)


#market cap

price_data$mcap <- price_data$Price * price_data$Shares.Outstanding * 1000

IDs <- unique(price_data$ID)
head(IDs)
mcap <- list()
for (i in 1:length(IDs)) {
  price_data_i <- price_data[price_data$ID==IDs[i], ]
  mcap[[i]] <- xts(price_data_i$mcap,
          order.by=as.Date(as.character(price_data_i$Date), format="%Y-%m-%d"))
}

mcap[1]

mcap <- do.call(cbind, mcap)
colnames(mcap) <- paste0("X", IDs)
dim(price)
mcap <- mcap["1990/"]
mcap[1:10,1:5]

mcap <- apply.daily(mcap, FUN=function(x) x[1,] )
mcap <- abs(mcap)
```

## Task 1.1 

```{r}
mcap_end_2002 <- mcap["2002-12-31"]

sorted_mcap__end_2002 <- order(mcap_end_2002, decreasing=TRUE)

largest_400 <- sorted_mcap__end_2002[1:400]

mcap_largest_400 <- mcap_end_2002[,largest_400]

price_end_2002 <- price["2002-12-31"]
price_end_2003 <- price["2003-12-31"]

price_end_2002_largest_400 <- price_end_2002[,largest_400]
price_end_2003_largest_400 <- price_end_2003[,largest_400]

returns_2003 <- (as.vector(price_end_2003_largest_400)/
                   as.vector(price_end_2002_largest_400))-1
plot(as.vector(mcap_largest_400),returns_2003,
     main="Returns of the largest 400 firms as a function of size",
     xlab="Market capitalization (2002)", ylab="Yearly return (2003)")

# exclude outliers
filtered_firms <-
  as.vector(mcap_largest_400) > quantile(as.vector(mcap_largest_400),0.05) &
  as.vector(mcap_largest_400) < quantile(as.vector(mcap_largest_400),0.95)

plot(as.vector(mcap_largest_400)[filtered_firms],
     returns_2003[filtered_firms],
    main="Returns as a function of size
    for firms between the 5% and 95% range of market cap",
    xlab="Market capitalization (2002)", ylab="Yearly return (2003)")
```

## Task 1.2 

```{r}
bin_breaks <- quantile(as.vector(mcap_largest_400), probs = seq(0, 1, by = 0.1))
mcap_deciles <- cut(as.vector(mcap_largest_400), breaks = bin_breaks, include.lowest = TRUE)
levels(mcap_deciles) <- paste0("q", 1:10)

returns_decile <- data.frame(returns_2003, mcap_deciles)

average_return_decile <- aggregate(returns_2003 ~ mcap_deciles, data = returns_decile, FUN = mean)
decile_avg_returns <- average_return_decile$returns_2003
names(decile_avg_returns) <- average_return_decile$mcap_deciles

months_2003 <- seq.Date(from = as.Date("2003-01-01"), to = as.Date("2003-12-31"), by = "month")

cumulative_returns <- xts(matrix(0, nrow = length(months_2003), ncol = length(decile_avg_returns)),
        order.by = months_2003)

colnames(cumulative_returns) <- names(decile_avg_returns)

for (d in 1:length(decile_avg_returns)) {
  avg_return <- decile_avg_returns[d]
  cumulative_returns[, d] <- cumprod(rep(1 + avg_return, length(months_2003))) - 1
}

plot(cumulative_returns[, 1], main = "Cumulative Monthly Returns by Decile", ylab = "Cumulative Return")  # First decile is used as the base plot

for (i in 2:10) {
  lines(cumulative_returns[, i], col=i)
}

addLegend("topleft", legend.names=paste0("Portfolio", 1:10), lwd=1, col=1:10)

average_yearly_returns <- decile_avg_returns * 12  # Annualized returns

barplot(average_yearly_returns, col = "lightblue", main = "Average yearly returns by decile",
        xlab = "Decile", ylab = "Average yearly return (2003)")
```
Observation: The average yearly return of the deciles follows a similar path
as the returns of the single firms.
However, they are less volatile and follow a smoother trend over time as there are less variablility. The outliers are not visible here due to the averaging effect of the deciles. This leads us to conclude the importance of diversification.

## Task 1.3

To test whether the stocks sorted by size have different returns, we use a paired t.test.
```{r}
r_q1 <- returns_decile$returns_2003[returns_decile$mcap_deciles == "q1"]
r_q10 <- returns_decile$returns_2003[returns_decile$mcap_deciles == "q10"]
r_q9 <- returns_decile$returns_2003[returns_decile$mcap_deciles == "q9"]
t.test(x=r_q1, y=r_q10, paired=TRUE, var.equal=FALSE)
```

We see that the p-value is larger than 0.05, 
therefore we fail to reject H0 and clonclude that there is no significant difference in returns between q1 and q10
```{r}
summary(r_q1)
summary(r_q10)
```
We see that the distributions are different.
We conduct another t.test for the returns of q9 and q1.
```{r}
t.test(x=r_q1, y=r_q9, paired=TRUE, var.equal =FALSE )
```
We see that the p-value is smaller than 0.05, 
therefore we reject H0 and clonclude that there is a significant difference in returns between q1 and q9

# Exercise 2

## Task 2.1
```{r, warnings=FALSE}
# Filter data to specific time range and downsample monthly
price <- na.locf(price)[endpoints(price, on="months")]
price <- price["2002-12-31/2018-12-31"]
price[1:10,1:5]
mcap <- mcap["2002-12-31/2018-12-31"]
mcap[1:10,1:5]

# Calculate monthly returns
monthly_returns <- (price / lag(price)) - 1

# Calculate yearly market cap and lag to previous year
mcap_year_end <- lag(mcap[endpoints(mcap, on = "years"), ])[-1, ]

# Compute yearly returns
yearly_returns <- apply.yearly(monthly_returns, function(ret) {
  apply(ret, 2, function(r) prod(1 + r) - 1)
})[-1, ]

# Remove extreme 1% of returns
return_quantiles <- quantile(yearly_returns, probs = c(0.005, 0.995), na.rm = TRUE)
yearly_returns[yearly_returns < return_quantiles[1] | yearly_returns > return_quantiles[2]] <- NA
```

## Task 2.2

```{r}
# Initialize an empty list for decile portfolio returns
decile_returns <- vector("list", length = nrow(yearly_returns))

# Iterate through each year to calculate deciles
for (year_idx in seq_len(nrow(yearly_returns))) {
  # Compute decile thresholds based on market capitalization
  decile_breaks <- quantile(mcap_year_end[year_idx, ], probs = seq(0, 1, 0.1), na.rm = TRUE)
  
  # Assign stocks to deciles
  decile_groups <- findInterval(coredata(mcap_year_end[year_idx, ]), decile_breaks, all.inside = TRUE)
  
  # Compute mean yearly returns for each decile
  decile_avg_returns <- tapply(
    coredata(yearly_returns[year_idx, ]),
    decile_groups,
    function(x) mean(x, na.rm = TRUE)
  )
  
  # Store the decile returns in the list
  decile_returns[[year_idx]] <- decile_avg_returns
}

# Combine decile returns into a matrix and convert to xts
decile_portfolio_matrix <- do.call(rbind, decile_returns)
colnames(decile_portfolio_matrix) <- paste0("Decile", seq_len(10))
rownames(decile_portfolio_matrix) <- rownames(yearly_returns)
decile_portfolio <- xts(decile_portfolio_matrix, order.by = index(yearly_returns))
head(decile_portfolio)
```

## Task 2.3 

```{r}

# Calculate cumulative portfolio returns
cumulative_returns <- cumprod(1 + decile_portfolio) - 1

# Define line colors for deciles

# Plot average yearly returns
average_yearly_plot <- plot(
  decile_portfolio,
  main = "Average Yearly Returns",
  ylim = c(-0.5, 1)
)
average_yearly_plot <- addLegend(
  legend.loc = "top",
  legend.names = paste0("d", 1:10),
  lty = 1,
  lwd = 2,
  horiz=TRUE,
  cex=0.65
)
average_yearly_plot

# Plot cumulative returns
cumulative_returns_plot <- plot(
  cumulative_returns,
  main = "Cumulative Return of the ten Portfolios"
)
cumulative_returns_plot <- addLegend(
  legend.loc = "topleft",
  legend.names = paste0("d", 1:10),
  lty = 1,
  lwd = 2,
)
cumulative_returns_plot

```


## Task 2.4

```{r}
# Extract returns for the smallest and largest decile portfolios
returns_smallest <- decile_portfolio[, "Decile1"]  # Smallest decile
returns_largest <- decile_portfolio[, "Decile10"] # Largest decile

# Remove NA values for clean comparison
clean_returns_smallest <- na.omit(returns_smallest)
clean_returns_largest <- na.omit(returns_largest)

# Perform Welch's Two-Sample t-test
t_test_result <- t.test(
  x = clean_returns_smallest,
  y = clean_returns_largest,
  alternative = "two.sided",
  var.equal = FALSE
)

# Display the t-test result
t_test_result
```

The t-test result indicates no significant difference in the mean returns between the smallest and largest decile portfolios at a 5% level of significance, as the p-value is relatively big. It does not seem that the size of the portfolio will make any significant difference in average returns within this dataset.
This result is supported by similar fluctuations shown in the graph of yearly returns throughout deciles.

