---
title: "Business Analytics II - Group 1 - Home Assignment 5"
author: "xxx xxx, xxx xxx, xxx xxx, xxx xxx, Benjamin Meixner, Nicholas Olsa"
date: "2024-12-16"
output: 
  pdf_document:
    latex_engine: tectonic
---
# Exercise 1
```{r}
# Load necessary libraries
library(lubridate)
library(xts)
cols <- c("steelblue", "coral", "forestgreen") # Colors for plots

# load stock returns
load("../data/returns_hw4.RData")
time(r) <- floor_date(time(r), unit="month") #round Date to 1st of current month (to match dates with risk-free rates below)


# load FF factors to get the risk-free rate
ff_factors <- read.csv(file = "../data/F-F_Research_Data_Factors.CSV")
head(ff_factors)
rf <- xts(ff_factors$RF/100, order.by=as.Date(paste0(ff_factors$X, "01"), format="%Y%m%d")) # Convert risk-free rate into xts object
colnames(rf) <- "rf"

rf <- rf[time(r)] # Subset risk-free rates to match the dates in the stock returns
rf


# Compute "equally weighted" market return
r_mkt <- xts(rowMeans(r), order.by=time(r)) # Average returns across all stocks for each time period
plot(cumprod(1+r_mkt)) # Plot cumulative market returns over time
```

Over 27 years we can see an accumulated returns of 35% with a significant dip in 2008.

```{r}

stock_picks <- c("X10026", "X10308", "X10443") # Define specific stock picks for analysis


# Rolling estimation of betas (10-year rolling window, updated yearly)
betas <- rollapply(cbind(r[,stock_picks], r_mkt)-c(coredata(rf)), width=10*12, by=12, fill=NULL, align="right", by.column=FALSE,
                   FUN=function(x){
                     beta <- rep(NA_real_, 3) # Initialize beta vector for three stocks
                     for(i in 1:3){
                       beta[i] <- coef(lm(x[,i]~x[,4]))[2]  # Calculate beta as slope of regression on market return
                     }
                     return(beta)
                   } )

tail(betas)  # Show the last computed betas
# betas = volatility of risk - how the returns of the stock are correlated to the overall market movement
plot(betas, col=cols) # Plot rolling betas for the selected stocks
```

Here the betas of our 3 picked stocks. We can see that in 2008 the blue line dips towards 1. We interpret that during the financial 
crisis investors are more risk-averse and therefore the risk is approaching market averages due to less risky buying decisions.

We can also see that the stocks we picked are in 2018 rather "low risk" since their volatility is only 0.8 while the market volatility would represent 1.

```{r}
# Rolling estimation of alphas (similar to betas)
alphas <- rollapply(cbind(r[,stock_picks], r_mkt)-c(coredata(rf)) , width=10*12, by=12, fill=NULL, align="right", by.column=FALSE,
                   FUN=function(x){
                     alpha <- rep(NA_real_, 3)  # Initialize alpha vector
                     for(i in 1:3){
                       alpha[i] <- coef(lm(x[,i]~x[,4]))[1]  # Calculate alpha as intercept of regression on market return
                     }
                     return(alpha)
                   } )

tail(alphas)
plot(alphas, col=cols) # Plot rolling alphas for the selected stocks
```

Here the alphas of our 3 picked stocks. We can see that the returns relative to the market returns are mostly higher on average for our portfolio, we would beat the
market with this arrangement. The blue stock has negative relative performance until 2005. The best stock is green while orange is close to the market average.

# Exercise 2
```{r}
# 1) COMPUTING THE BETAS OF ALL STOCKS UP TO 2006

# Create regression matrix X: first column is intercept, second is excess market returns
# -> used to compute linear regression coefficients as (x'X)^-1 X'y
X <- cbind(1, coredata(r_mkt["/2006"] - rf["/2006"]))

# (x'X)^-1 X'y to obtain linear regression coefficients
coefs <- solve(t(X)%*%X) %*% t(X) %*% coredata(r["/2006"] - as.vector(rf["/2006"])) #first row contains alphas, second row betas, columns correspond to firms

betas <- coefs[2,]  # Extract betas from the second row

# Identify top 3 stocks with lowest and highest betas
low_beta <- head(sort(betas), 3) # Lowest 3 betas
high_beta <- tail(sort(betas), 3)  # Highest 3 betas


# Compute cumulative returns (2007-2010) for low and high beta stocks
r_low_beta <- apply(r[,names(low_beta)]["2007/2010"], 2, FUN=\(x) cumprod(1+x) )
# Compute cumulative product of returns for low-beta stocks
r_low_beta <- xts(r_low_beta, order.by=ymd(rownames(r_low_beta)) )

r_high_beta <- apply(r[,names(high_beta)]["2007/2010"], 2, FUN=\(x) cumprod(1+x) )
# Compute cumulative product of returns for high-beta stocks
r_high_beta <- xts(r_high_beta, order.by=ymd(rownames(r_high_beta)) )

# Plot cumulative returns
par(mfrow=c(1,2))
plot(r_low_beta)
plot(r_high_beta)
par(mfrow=c(1,1))
```
Here we can clearly see how the beta is a "multiplier of market movement". While the low betas are mostly stable for themselves and ignore the financial crisis
the high beta stocks are following the market considerably - up as well as down.

# Exercise 3
```{r}
## load the data
ff_factors <- read.csv(file = "../data/F-F_Research_Data_Factors.CSV")
industries<-read.csv(file = "../data/12_Industry_Portfolios.CSV")
head(ff_factors)
head(industries)

all(ff_factors$X == industries$X) # Check that all dates are the same

## Calculate industry excess returns
industries_xret <- cbind(industries[,1], industries[,-1]-ff_factors$RF)

# Combine industry excess returns and Fama-French factors into one dataset
xret_data <- cbind(industries_xret, ff_factors[,-1])

# Rolling regressions over two-year windows (24 months)
betas <- rollapply(data = xret_data, width = 24, by.column = FALSE, FUN = function(x){
  #print((x))
  b <- list() # Store regression results for each industry
  y <- as.matrix(x[,2:13]) # Dependent variables: industry excess returns
  X <- cbind(rep(1,24), as.matrix(x[,14:16])) # Independent variables: intercept, market, SMB, HML returns
  for(i in 1:12) b[[i]] <- solve(t(X)%*%X)%*%t(X)%*%y[,i]  # Compute regression coefficients
  return(t(do.call(rbind,b))) # Combine results into a single matrix
})

alphas <- betas[,seq(1,48,4)] # Extract alphas (intercepts) for each industry
betas <- betas[,-seq(1,48,4)] # Remove intercepts, leaving only betas (market, SMB, HML)

# Rename columns for clarity
colnames(betas) <- paste0(rep(colnames(xret_data)[2:13],each=3), colnames(betas))
colnames(alphas) <- colnames(xret_data)[2:13]
# CRAN version
# install.packages("ggthemes")
library(ggthemes)
# Define time axis for plotting
xaxis_time <- as.yearmon(as.Date(paste(ff_factors$X,"01"), "%Y%m%d"))[-(1:23)]

par(mfrow=(c(1,3)))
## plot market betas
plot(xaxis_time, betas[,1], type="l", col=unlist(ggthemes_data$economist$fg[2])[1],
     ylim = c(min(betas[,seq(1,36,3)]), max(betas[,seq(1,36,3)])),
     ylab="Market Beta", xlab="time")
for(i in seq(4,36,3)) lines(xaxis_time, betas[,i], col=unlist(ggthemes_data$economist$fg[2])[ceiling(i/3)])

## plot smb betas
plot(xaxis_time,betas[,2], type="l", col=unlist(ggthemes_data$economist$fg[2])[1],
     ylim = c(min(betas[,seq(2,36,3)]), max(betas[,seq(2,36,3)])),
     ylab="SMB Beta", xlab="time")
for(i in seq(5,36,3)) lines(xaxis_time, betas[,i], col=unlist(ggthemes_data$economist$fg[2])[ceiling(i/3)])


## plot hml betas
plot(xaxis_time,betas[,3], type="l", col=unlist(ggthemes_data$economist$fg[2])[1],
     ylim = c(min(betas[,seq(3,36,3)]), max(betas[,seq(3,36,3)])),
     ylab="HML Beta", xlab="time")
for(i in seq(6,36,3)) lines(xaxis_time, betas[,i], col=unlist(ggthemes_data$economist$fg[2])[ceiling(i/3)])
par(mfrow=c(1,1))
```

Here the 3 different betas of the Fama-French model.

Taking the Market beta as a baseline we can see that the SMB beta is centered around 0 rather than 1 - low-cap stocks are much more volatile.

Turning to HML beta we can see that the spread is larger for the growth stocks - they have higher extremes for the expense of a center around 1.
Here we can be truly market-beating with the right growth stocks, but finding the right stock(s) is hard.

```{r}
## plot average betas
barplot(height = t(matrix(colMeans(betas), ncol=3, byrow=TRUE)), beside=TRUE,
        col = unlist(ggthemes_data$economist$fg[2])[1:3], border = FALSE, 
        names.arg = (names(industries)[-1]), las=2)
par(xpd=TRUE)
legend(x = -0.7, y = 1.4, legend = c("market", "smb", "hml"),
       col=unlist(ggthemes_data$economist$fg[2])[1:3], pch=15, bty="n", border=FALSE)
par(xpd=FALSE)      
```

Here we can see that the market beta is relatively uniform. Growth stocks are oppositely market-following in Business Equipment and Health, optimal picks for 
uncertain times while in Energy, Chemistry, Telecomunications and Utilities the small cap stocks are moving against the market.

```{r}
## plot alphas over time
plot(xaxis_time, alphas[,1], type="l", col=unlist(ggthemes_data$economist$fg[2])[1],
     ylim = c(min(alphas), max(alphas)),
     ylab="Alpha", xlab="time")
for(i in 2:12) lines(xaxis_time, alphas[,i], col=unlist(ggthemes_data$economist$fg[2])[i])
```

The alphas over time average at about 0 - market returns. Owning all stocks returns market returns - a revelation!

```{r}
# Plot average alphas for each industry
barplot(height = colMeans(alphas), 
        col = unlist(ggthemes_data$economist$fg[2])[1:12], border = FALSE, 
        names.arg = (names(industries)[-1]), las=2, main = "Industry Alphas")
```

Here we can see which sectors produce most market-beating returns. Health, Business Equiment and Telecommunications seem like 
the have good marketing-beating potential while "Other" sectors produce lower than market returns.


