---
title: "HA3_tasks1,2"
output:
  pdf_document: 
    latex_engine: tectonic
    fig_caption: yes
authors: "Benjamin Meixner, Maryna Manzharova, Elias Clemens Maria Prackwieser, Richard Takacs"
date: "`r Sys.Date()`"
---

```{r}
library(xts)
library(lubridate)
```

# Exercise 1 

First, we call part of the code from class 3 that we will reuse for the exercises. 

```{r}
price_data <- read.csv(file="../data/price_data.csv", sep = ";", dec = ",")
firm_data <- read.csv(file="../data/firm_data.csv", sep = ";", dec = ",")

price_data <- na.omit(price_data)
firm_data <- na.omit(firm_data)

IDs <- unique(price_data$ID)

# Creating xts object for prices of all firms
IDs <- unique(price_data$ID)
head(IDs)
p <- list()
for (i in 1:length(IDs)) { 
  price_data_i <- price_data[price_data$ID==IDs[i], ]
  p[[i]] <- xts(price_data_i$Price,
       order.by=as.Date(as.character(price_data_i$Date), format="%Y-%m-%d"))
}
price <- do.call(cbind, p)
colnames(price) <- paste0("X", IDs)
dim(price)
price <- price["1990/"]
price[1:10,1:5]

price <- apply.daily(price, FUN=function(x) x[1,] )
dim(price)
price <- abs(price)
price <- na.approx(price)

#market cap

price_data$mcap <- price_data$Price * price_data$Shares.Outstanding * 1000

IDs <- unique(price_data$ID)
head(IDs)
mcap <- list()
for (i in 1:length(IDs)) {
  price_data_i <- price_data[price_data$ID==IDs[i], ]
  mcap[[i]] <- xts(price_data_i$mcap,
          order.by=as.Date(as.character(price_data_i$Date), format="%Y-%m-%d"))
}

mcap[1]

mcap <- do.call(cbind, mcap)
colnames(mcap) <- paste0("X", IDs)
dim(price)
mcap <- mcap["1990/"]
mcap[1:10,1:5]

mcap <- apply.daily(mcap, FUN=function(x) x[1,] )
mcap <- abs(mcap)
```
## Task 1 
```{r}
mcap_end_2002 <- mcap["2002-12-31"]

#Firms are sorted by market cap in descending order
sorted_mcap__end_2002 <- order(mcap_end_2002, decreasing=TRUE)

#We chose to work with the largest 400 firms
largest_400 <- sorted_mcap__end_2002[1:400]

#Subsetting market cap to include only our selected firms
mcap_largest_400 <- mcap_end_2002[,largest_400]

price_end_2002 <- price["2002-12-31"]
price_end_2003 <- price["2003-12-31"]

#subsetting year end prices for the 400 largest firms
price_end_2002_largest_400 <- price_end_2002[,largest_400]
price_end_2003_largest_400 <- price_end_2003[,largest_400]

#simple return for 2003 of the 400 largest firms
returns_2003 <- (as.vector(price_end_2003_largest_400)/
                   as.vector(price_end_2002_largest_400))-1
plot(as.vector(mcap_largest_400),returns_2003,
     main="Returns of the largest 400 firms as a function of size",
     xlab="Market capitalization (2002)", ylab="Yearly return (2003)")

#only firms between 5% and 95% range of market cap --> excluding outliers
filtered_firms <-
  as.vector(mcap_largest_400) >quantile(as.vector(mcap_largest_400),0.05) &
  as.vector(mcap_largest_400) < quantile(as.vector(mcap_largest_400),0.95)

plot(as.vector(mcap_largest_400)[filtered_firms],
     returns_2003[filtered_firms],
    main="Returns as a function of size
    for firms between the 5% and 95% range of market cap",
    xlab="Market capitalization (2002)", ylab="Yearly return (2003)")
```

## Task 2 

```{r}
bin_breaks <- quantile(as.vector(mcap_largest_400), probs =seq(0,1,by=0.1))
mcap_deciles <- cut(as.vector(mcap_largest_400),breaks=bin_breaks, include.lowest=TRUE)
levels(mcap_deciles) <- paste0("q", 1:10)
mcap_deciles

#Combining returns and deciles into a data frame
returns_decile <- data.frame(returns_2003, mcap_deciles)

#average return per each decile
average_return_decile <- aggregate(returns_2003~mcap_deciles, data=returns_decile, FUN=mean)

#Specifying the two columns of the data frame
decile_avg_returns <- average_return_decile$returns_2003
names(decile_avg_returns) <- average_return_decile$mcap_deciles

#Creating a time sequence for 12 months of 2003
months_2003 <- seq.Date(from = as.Date("2003-01-01"), to = as.Date("2003-12-31"), by = "month")

#Initializing cumulative returns for each decile with a matrix filled with 0-s
cumulative_returns <- xts(matrix(0, nrow = length(months_2003), ncol = length(decile_avg_returns)),
                          order.by = months_2003)
colnames(cumulative_returns) <- names(decile_avg_returns)

#Computing cumulative monthly returns for each decile using a loop
for (d in 1:length(decile_avg_returns)) {
  avg_return <- decile_avg_returns[d]
  cumulative_returns[, d] <- cumprod(rep(1 + avg_return, length(months_2003))) - 1
}

#Plotting cumulative returns
plot(index(cumulative_returns), cumulative_returns[, 1], type = "l",
     main = "Cumulative Monthly Returns by Decile", xlab = "Time", ylab = "Cumulative Return")  # First decile is used as the base plot

for (d in 2:ncol(cumulative_returns)) {
  lines(index(cumulative_returns), cumulative_returns[, d], col = d, lwd = 2) 
  #other deciles are added with the loop
}

legend("topright", legend = colnames(cumulative_returns),
       col = 1:ncol(cumulative_returns), lty = 1, lwd = 2)

#Converting yearly decile returns to average yearly returns
average_yearly_returns <- decile_avg_returns * 12  # Annualized returns

#Plotting bar chart of yearly returns
barplot(average_yearly_returns, col = "lightblue", main = "Average yearly returns by decile",
        xlab = "Decile", ylab = "Average yearly return (2003)")
```
Observation: The average yearly return of the deciles follows a similar path
as the returns of the single firms.
However, they are less volatile and follow a smoother trend over time as there are less variablility. The outliers are not visible here due to the averaging effect of the deciles. This leads us to conclude the importance of diversification.

# Exercise 2

## 2.1

```{r}
# Initialize lists to store yearly returns and market caps
yearly_returns <- list()
yearly_mcap <- list()

# Loop through each year from 2002 to 2018
for (year in 2002:2018) {
  # Get market cap and prices at the end of the current year and the next year
  mcap_end_year <- na.omit(mcap[paste0(year, "-12-31")])
  price_end_year <- na.omit(price[paste0(year, "-12-31")])
  price_end_next_year <- na.omit(price[paste0(year + 1, "-12-31")])
  
  # Calculate returns for the next year
  returns_next_year <- (as.vector(price_end_next_year) / as.vector(price_end_year)) - 1
  
  # Set the most extreme 1% of returns to NA
  returns_next_year[returns_next_year < quantile(returns_next_year, 0.005) | 
                    returns_next_year > quantile(returns_next_year, 0.995)] <- NA
  
  # Store the returns and market caps
  yearly_returns[[as.character(year)]] <- returns_next_year
  yearly_mcap[[as.character(year)]] <- as.vector(mcap_end_year)
}
```

## 2.2

```{r}
# Initialize a matrix to store cumulative returns for each decile
cumulative_returns_deciles <- matrix(0, nrow = length(2002:2018), ncol = 10)
rownames(cumulative_returns_deciles) <- 2002:2018
colnames(cumulative_returns_deciles) <- paste0("Decile ", 1:10)

# Loop through each year to compute decile returns
for (year in 2002:2018) {
  # Get the returns and market caps for the current year
  returns <- yearly_returns[[as.character(year)]]
  mcap <- yearly_mcap[[as.character(year)]]
  
  # Remove NA values
  valid_indices <- !is.na(returns) & !is.na(mcap)
  returns <- returns[valid_indices]
  mcap <- mcap[valid_indices]
  
  # Classify firms into deciles based on market cap
  bin_breaks <- quantile(mcap, probs = seq(0, 1, by = 0.1))
  mcap_deciles <- cut(mcap, breaks = bin_breaks, include.lowest = TRUE)
  levels(mcap_deciles) <- paste0("Decile ", 1:10)
  
  # Compute average returns for each decile
  returns_decile <- data.frame(returns, mcap_deciles)
  average_return_decile <- aggregate(returns ~ mcap_deciles, data = returns_decile, FUN = mean)
  
  # Store the average returns in the cumulative returns matrix
  cumulative_returns_deciles[as.character(year), ] <- average_return_decile$returns
}
```

```{r}
# Compute cumulative returns for each decile
cumulative_returns_deciles <- apply(cumulative_returns_deciles + 1, 2, cumprod) - 1

# Plot cumulative returns for each decile
plot(2002:2018, cumulative_returns_deciles[, 1], type = "l", col = 1, lwd = 2,
     main = "Cumulative Returns by Decile (2002-2018)", xlab = "Year", ylab = "Cumulative Return")
for (d in 2:10) {
  lines(2002:2018, cumulative_returns_deciles[, d], col = d, lwd = 2)
}
legend("topright", legend = colnames(cumulative_returns_deciles), col = 1:10, lty = 1, lwd = 2)
```

```{r}
# Compute average yearly returns for each decile
average_yearly_returns_deciles <- colMeans(cumulative_returns_deciles)

# Plot bar chart of average yearly returns
barplot(average_yearly_returns_deciles, col = "lightblue", main = "Average Yearly Returns by Decile (2002-2018)",
        xlab = "Decile", ylab = "Average Yearly Return")
```

```{r}
# Perform parametric hypothesis test to check for differences in returns
anova_result <- aov(returns ~ mcap_deciles, data = returns_decile)
summary(anova_result)
```

