---
title: "Homework Unit 6 BA1"
output: 
  pdf_document:
    latex_engine: tectonic
date: "2024-12-15"
author: "Benjamin Meixner, Maryna Manzharova, Elias Clemens Maria Prackwieser, Richard Takacs"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

``` {r}
library(readxl)
library(Benchmarking)

setwd("~/Downloads") #set the working directory where you have the input file !!!!

# read the input data file
df <- data.frame(read_excel(path = "Data-Assignment.xlsx", sheet="Sheet1"))

df_15 <-df[15,]  # for #15 
# inputs and outputs to be used for deriving targets 
inputs=data.frame(df[1:2]) # 1:2 - columns where the inputs are entered
outputs=data.frame(df[3:6]) # 3:6 - columns where the outputs are entered

# DEA analysis
result<-dea(inputs, outputs, RTS="crs", ORIENTATION="in", XREF=NULL, YREF=NULL,FRONT.IDX=NULL, SLACK=TRUE, DUAL=FALSE, DIRECT=NULL, param=NULL,TRANSPOSE=FALSE, FAST=FALSE, LP=FALSE, CONTROL=NULL, LPK=NULL)

# create a data frame with the result and write it on a .csv file
N = nrow(df) # number of DMUs
Solution = data.frame(seq(1,N), result$eff,result$sx,result$sy) #result$eff: efficiency scores, result$sx: input excess, result$sy: output shortfall
colnames(Solution) = c('DMU','effi',colnames(df)) # assign column names
write.csv(Solution, "Solution-Example.csv")


result_sol <- read.csv("Solution-Example.csv")
result_sol

```
```{r}
result_DMU15<- result_sol[15,]
result_DMU15
eff_DMU15 <- result_DMU15$effi
eff_DMU15 #efficiency score for DMU 15
``` 

We can see that for DMU 15 efficiency score is 0.9423003, which is below 1. It has also shortage in outputs Economic Perfomance, Operational Perfomance and Social Perfomance.
The DMU is not efficient. Adjustment have to be made.
At first, inputs must be reduced to $Efficiency\ Score \cdot x_{i}$
```{r}
New_Lean <- df_15$Lean.Practice  * eff_DMU15
New_Lean
New_innov <- df_15$Innovation  * eff_DMU15
New_innov
```
Then, we have to adjust outputs that have shortage (add them to inputs)
We have to add  0.6219242 to Economic Performance output.
We have to add  0.005530871 to Operational Performance output.
We have to add  0.004178742 to Social Performance output.
``` {r}
New_Ec <- df_15$Economic.Performance + result_DMU15$Economic.Performance
New_Ec
New_Oper <- df_15$Operational.Performance + result_DMU15$Operational.Performance
New_Oper
New_Social <- df_15$Social.Performance + result_DMU15$Social.Performance
New_Social
```
```{r}
df[15,]<- c(New_Lean, New_innov, New_Ec, New_Oper, df_15$Environment.Performance, New_Social)
df[15,]
```

#DEA analysis 
``` {r}
inputs=data.frame(df[1:2]) # 1:2 - columns where the inputs are entered
outputs=data.frame(df[3:6]) # 3:6 - columns where the outputs are entered

result<-dea(inputs, outputs, RTS="crs", ORIENTATION="in", XREF=NULL, YREF=NULL,FRONT.IDX=NULL, SLACK=TRUE, DUAL=FALSE, DIRECT=NULL, param=NULL,TRANSPOSE=FALSE, FAST=FALSE, LP=FALSE, CONTROL=NULL, LPK=NULL)

N = nrow(df) # number of DMUs
Solution = data.frame(seq(1,N), result$eff,result$sx,result$sy) #result$eff: efficiency scores, result$sx: input excess, result$sy: output shortfall
colnames(Solution) = c('DMU','effi',colnames(df)) # assign column names
write.csv(Solution, "Solution-Example.csv")
result_sol <- read.csv("Solution-Example.csv")
result_DMU15<- result_sol[15,]
result_DMU15 # now, DMU 15 is efficient. 
```
