---
title: "Business Analytics II - Group 1 - Home Assignment 4"
author: "Yurii Masiuk, Vanda Lakatos, Kornel Auer, Vivien Arns, Benjamin Meixner, Nicholas Olsa"
date: "2024-12-04"
output: 
  pdf_document:
    latex_engine: tectonic
---
# Exercise 1
```{r}
library(xts)

## Plot the PF frontier in the two asset case with the following parameters
## E(rA)=.2, E(rB)=.5, sd(rA)=.2, sd(rB)=.4, rho(rA,rB)=-.2

## We start by writing a function that computes the frontier for different input values
frontier <- function(mu.pf, mu, s, rho){
  return(sqrt( (1/(mu[2]-mu[1])^2) * 
                 ( (mu.pf-mu[2])^2 * s[1]^2 + (mu.pf-mu[1])^2 * s[2]^2 -
                     2*rho*s[1]*s[2]*(mu.pf-mu[1])*(mu.pf-mu[2])))  )
}

pf.return <- function(mu, w) return(t(w)%*%mu)
test_pf_returns <- seq(-1,1,.01)


plot(frontier(test_pf_returns, mu = c(.2,.5), s = c(.2,.4), rho=-0.2), test_pf_returns, 
     ylab="Exp. PF return", xlab="Risk", type="l", lwd=2,
     main="PF Frontier for different rho", ylim = c(-1,1), xlim = c(0,3))
lines(frontier(test_pf_returns, 
               mu = c(.2,.5), 
               s = c(.2,.4), 
               rho=0), 
      test_pf_returns, col="blue", lwd=2, lty=2)
lines(frontier(test_pf_returns, 
               mu = c(.2,.5), 
               s = c(.2,.4), 
               rho=0.8), 
      test_pf_returns, col="red", lwd=2, lty=3)
lines(frontier(test_pf_returns, 
               mu = c(.2,.5), 
               s = c(.2,.4), 
               rho=-0.8), 
      test_pf_returns, col="purple", lwd=2, lty=4)
legend("topright", legend=c(-0.8, -0.2, 0, 0.8), title="Rho", bty="n",
       col=c("purple", "black", "blue", "red"), lwd=2, lty=c(4, 1, 2, 3))
```

Lower or negative correlation increases the curvature of the portfolio frontier, which translates to better diversification benefits. This allows portfolios to achieve lower risk for the same expected return or higher returns for the same level of risk.With high positive correlation (rho = 0.8), diversification benefits are almost nonexistent, as both assets tend to move together. In practical terms that means Investors aiming to minimize risk should target portfolios along the most curved frontier

# Exercise 2
```{r}
## We start by coding a function that computes the frontier with/without a short 

ba2_optimal_portfolio<-function(asset.eret, asset.cov, 
                                no.short_selling=TRUE, max.weight=NULL, min.weight=NULL,
                                frontier.r=seq(min(asset.eret),max(asset.eret),0.001),
                                fully.invested=TRUE){
  require(quadprog) # check if package 'quadprog' is loaded; if not, load it
  
  k <- length(asset.eret) # get number of stocks
  dvec <- rep(0,k) # create zero vector 
  
  ## Code the constraints
  if(fully.invested){
    c1 <- rep(1,k) # weights need to sum to one
    eq_constraints <- 2
    c2 <- asset.eret # weighted expected returns need to be expected PF return
    C <- t(rbind(c1,c2)) # this matrix represents the LHS of all constraints
  }else{
    c2 <- asset.eret # weighted expected returns need to be expected PF return
    eq_constraints <- 1
    C <- matrix(c2,ncol=1) # this matrix represents the LHS of all constraints
  }
  

  
  ## Short selling constraint
  if(no.short_selling){
    c3 <- matrix(0, ncol=k, nrow=k)
    diag(c3) <- 1
    C <- t(rbind(t(C),c3))
  }
  
  ## Maximum weight per asset
  if(!is.null(max.weight)){
    c4 <- matrix(0, ncol=k, nrow=k)
    diag(c4) <- -1 # since it is a geq constraints we check if (-1)*w >= -max.weight
    C <- t(rbind(t(C),c4))
  }
  
  ## Minimum weight per asset
  if(!is.null(min.weight)){
    c5 <- matrix(0, ncol=k, nrow=k)
    diag(c5) <- 1 # since it is a geq constraints we check if w >= min.weight
    C <- t(rbind(t(C),c5))
  }
  
  ## Create vector that contains the estimated min variance given an 
  ## expected portfolio return target
  frontier.s <- rep(NA_real_, length(frontier.r))
  ## Create a container for the corresponding portfolio weights
  pf.weights <- list()
  
  for(i in 1:length(frontier.r)){
    
    if(fully.invested) c0 <- c(1, frontier.r[i]) #this vector contains the RHS of constraints
    else c0 <- frontier.r[i] #this vector contains the RHS of constraints
    if(no.short_selling) c0 <- c(c0, rep(0,k))
    if(!is.null(max.weight)) c0 <- c(c0, -rep(max.weight,k))
    if(!is.null(min.weight)) c0 <- c(c0, -rep(min.weight,k))
    
    w.solution <- solve.QP(Dmat=asset.cov, dvec=dvec, Amat=C, bvec=c0, meq=eq_constraints)
    ## We are interested in the standard deviation but are actually minimizing
    ## (1/2) of the variance, so we need to account for that!
    frontier.s[i] <- sqrt(w.solution$value*2) 
    pf.weights[[i]] <- w.solution$solution
  }
  
  return(list("Volatility"=frontier.s, "Optimal.Weights"=pf.weights,"Expected.Return"=frontier.r))
  
}


# function to compute portfolio return for given asset returns and portfolio weights
pf.return <- function(mu, w) return(t(w)%*%mu)

# function to compute portfolio vola for given asset return covariances and portfolio weights
pf.volatility <- function(Sigma, w) return(sqrt((t(w)%*%Sigma%*%w)))


## Load the return data set
load("../data/returns_hw4.RData")

# 1) Compute the frontier using returns up to 2001 (no short selling, no rf asset)

r_2001 <- r["/2001"] #subset return
expret_2001 <- colMeans(r_2001) # compute the expeceted returns
Sigma_2001 <- cov(r_2001) # compute the variance-covariance matrix

frontier_2001 <- ba2_optimal_portfolio(asset.eret = expret_2001, 
                                       asset.cov = Sigma_2001,
                                       no.short_selling = TRUE)

## 2) Compute the frontier using all returns (no short selling, no rf asset)
expret_full <- colMeans(r) # compute the expected returns
Sigma_full <- cov(r) # compute the variance-covariance matrix

frontier_full <- ba2_optimal_portfolio(asset.eret = expret_full, asset.cov = Sigma_full,
                                       no.short_selling = TRUE)

## Pick five portfolios (randomly)
set.seed(1234) # set seed to make random draw reproducible
five_pfs <- sample(length(frontier_2001$Optimal.Weights), size=5)
five_pfs

## Plot the result
par(mfrow=c(1,2))
plot(sqrt(diag(Sigma_2001)), expret_2001, pch=19, col="yellow", main="Until 2001", 
     ylim = c(-.02,.05), xlim=c(0,.3), ylab="Exp. Return", xlab="Risk")
lines(frontier_2001$Volatility, frontier_2001$Expected.Return, lwd=2, col="coral")
points(frontier_2001$Volatility[five_pfs], frontier_2001$Expected.Return[five_pfs], col="steelblue", pch=19)

plot(sqrt(diag(Sigma_full)), expret_full, pch=19, col="yellow", main="Full Sample", 
     ylim = c(-.02,.05), xlim=c(0,.3), ylab="Exp. Return", xlab="Risk")
lines(frontier_full$Volatility, frontier_full$Expected.Return, lwd=2, col="coral")
points(pf.volatility(Sigma_full, frontier_2001$Optimal.Weights[[five_pfs[1]]]), 
       pf.return(expret_full, frontier_2001$Optimal.Weights[[five_pfs[1]]]), 
       col="steelblue", pch=19)
points(pf.volatility(Sigma_full, frontier_2001$Optimal.Weights[[five_pfs[2]]]), 
       pf.return(expret_full, frontier_2001$Optimal.Weights[[five_pfs[2]]]), 
       col="steelblue", pch=19)
points(pf.volatility(Sigma_full, frontier_2001$Optimal.Weights[[five_pfs[3]]]), 
       pf.return(expret_full, frontier_2001$Optimal.Weights[[five_pfs[3]]]), 
       col="steelblue", pch=19)
points(pf.volatility(Sigma_full, frontier_2001$Optimal.Weights[[five_pfs[4]]]), 
       pf.return(expret_full, frontier_2001$Optimal.Weights[[five_pfs[4]]]), 
       col="steelblue", pch=19)
points(pf.volatility(Sigma_full, frontier_2001$Optimal.Weights[[five_pfs[5]]]), 
       pf.return(expret_full, frontier_2001$Optimal.Weights[[five_pfs[5]]]), 
       col="steelblue", pch=19)
par(mfrow=c(1,1))
```
We can observe that the full sample has less stread in the expected returns.
Also the left-most point where the portfolio is minimizing risk with the highest
respective expected return is higher if we only consider options until 2001.
Reading from the graphs we can see ~1% expected return from 'until 2001' and 
0.75% from the full sample.

# Exercise 3

```{r}
## We start by computing the frontier up until 2001
r_2001 <- r["/2001"] #subset return
expret_2001 <- colMeans(r_2001) # compute the expeceted returns
Sigma_2001 <- cov(r_2001) # compute the variance-covariance matrix

frontier_2001 <- ba2_optimal_portfolio(asset.eret = expret_2001, 
                                       asset.cov = Sigma_2001,
                                       no.short_selling = FALSE,
                                       # we add/subtract a small constant to avoid numerical 
                                       # instability precisely at the min/max asset return
                                       frontier.r = seq(min(expret_2001)+0.0001, 
                                                        max(expret_2001)-0.0001, 
                                                        0.001)) 

r_2002 <- r["/2002"] #subset return
expret_2002 <- colMeans(r_2002) # compute the expected returns
Sigma_2002 <- cov(r_2002) # compute the variance-covariance matrix

frontier_2002 <- ba2_optimal_portfolio(asset.eret = expret_2002, 
                                       asset.cov = Sigma_2002,
                                       no.short_selling = FALSE,
                                       frontier.r = seq(min(expret_2002)+0.0001, 
                                                        max(expret_2002)-0.0001, 
                                                        0.001))


three_pfs2001 <- c(which.min((frontier_2001$Expected.Return-0.01)^2),
               which.min((frontier_2001$Expected.Return-0.015)^2),
               which.min((frontier_2001$Expected.Return-0.02)^2))
three_pfs2002 <- c(which.min((frontier_2002$Expected.Return-0.01)^2),
                   which.min((frontier_2002$Expected.Return-0.015)^2),
                   which.min((frontier_2002$Expected.Return-0.02)^2))

par(mfrow=c(1,2))
plot(sqrt(diag(Sigma_2001)), expret_2001, pch=19, col="yellow", main="Until 2001", ylim = c(-.02,.05), xlim=c(0,.3), ylab="Exp. Return", xlab="Risk")
lines(frontier_2001$Volatility, frontier_2001$Expected.Return, lwd=2, col="coral")
points(frontier_2001$Volatility[three_pfs2001[1]], 
       frontier_2001$Expected.Return[three_pfs2001[1]], 
       col="steelblue", pch=19)
points(frontier_2001$Volatility[three_pfs2001[2]], 
       frontier_2001$Expected.Return[three_pfs2001[2]], 
       col="steelblue", pch=19)
points(frontier_2001$Volatility[three_pfs2001[3]], 
       frontier_2001$Expected.Return[three_pfs2001[3]], 
       col="steelblue", pch=19)

plot(sqrt(diag(Sigma_2002)), expret_2002, pch=19, col="yellow", main="Until 2002", ylim = c(-.02,.05), xlim=c(0,.3), ylab="Exp. Return", xlab="Risk")
lines(frontier_2002$Volatility, frontier_2002$Expected.Return, lwd=2, col="coral")
points(frontier_2002$Volatility[three_pfs2002[1]], 
       frontier_2002$Expected.Return[three_pfs2002[1]], 
       col="steelblue", pch=19)
points(frontier_2002$Volatility[three_pfs2002[2]], 
       frontier_2002$Expected.Return[three_pfs2002[2]], 
       col="steelblue", pch=19)
points(frontier_2002$Volatility[three_pfs2002[3]], 
       frontier_2002$Expected.Return[three_pfs2002[3]], 
       col="steelblue", pch=19)
par(mfrow=c(1,1))

```
The difference short selling makes on the portfolios is apparent right away. The risk 
is almost 0 while the expected are higher too with 1% - 2%.

```{r}
par(mfrow=c(2,3))
barplot(frontier_2001$Optimal.Weights[[three_pfs2001[1]]], 
        border=FALSE, col="black", main="1%")
barplot(frontier_2001$Optimal.Weights[[three_pfs2001[2]]], 
        border=FALSE, col="black", main="1.5%")
barplot(frontier_2001$Optimal.Weights[[three_pfs2001[3]]], 
        border=FALSE, col="black", main="2%")
barplot(frontier_2002$Optimal.Weights[[three_pfs2002[1]]], 
        border=FALSE, col="black", main="1%")
barplot(frontier_2002$Optimal.Weights[[three_pfs2002[2]]], 
        border=FALSE, col="black", main="1.5%")
barplot(frontier_2002$Optimal.Weights[[three_pfs2002[3]]], 
        border=FALSE, col="black", main="2%")
par(mfrow=c(1,1))

```
When plotting the weights of the portfolio we can clearly see that some stocks are bought
or short-sold differently based on the expected return and therefore different risk levels.
A line in the negative region of the plot means the stock is short-sold.
```{r}
weight_change <- c(mean(abs(frontier_2001$Optimal.Weights[[three_pfs2001[1]]]
                            -frontier_2002$Optimal.Weights[[three_pfs2002[1]]])),
                   mean(abs(frontier_2001$Optimal.Weights[[three_pfs2001[2]]]
                            -frontier_2002$Optimal.Weights[[three_pfs2002[2]]])),
                   mean(abs(frontier_2001$Optimal.Weights[[three_pfs2001[3]]]
                            -frontier_2002$Optimal.Weights[[three_pfs2002[3]]])))

## No short selling PFs
frontier_2001_ns <- ba2_optimal_portfolio(asset.eret = expret_2001, 
                                          asset.cov = Sigma_2001,
                                          no.short_selling = TRUE,
                                          frontier.r = seq(min(expret_2001)+0.0001, 
                                                           max(expret_2001)-0.0001, 
                                                           0.001))
frontier_2002_ns <- ba2_optimal_portfolio(asset.eret = expret_2002, 
                                          asset.cov = Sigma_2002,
                                          no.short_selling = TRUE,
                                          frontier.r = seq(min(expret_2002)+0.0001, 
                                                           max(expret_2002)-0.0001, 
                                                           0.001))
par(mfrow=c(1,2))
plot(sqrt(diag(Sigma_2001)), expret_2001, pch=19, col="yellow", 
     main="Until 2001", ylim = c(-.02,.05), xlim=c(0,.3), ylab="Exp. Return", xlab="Risk")
lines(frontier_2001_ns$Volatility, frontier_2001_ns$Expected.Return, lwd=2, col="coral")
points(frontier_2001_ns$Volatility[three_pfs2001[1]], 
       frontier_2001_ns$Expected.Return[three_pfs2001[1]], 
       col="steelblue", pch=19)
points(frontier_2001_ns$Volatility[three_pfs2001[2]], 
       frontier_2001_ns$Expected.Return[three_pfs2001[2]], 
       col="steelblue", pch=19)
points(frontier_2001_ns$Volatility[three_pfs2001[3]], 
       frontier_2001_ns$Expected.Return[three_pfs2001[3]], 
       col="steelblue", pch=19)

plot(sqrt(diag(Sigma_2002)), expret_2002, pch=19, col="yellow", 
     main="Until 2002", ylim = c(-.02,.05), xlim=c(0,.3), ylab="Exp. Return", xlab="Risk")
lines(frontier_2002_ns$Volatility, frontier_2002_ns$Expected.Return, lwd=2, col="coral")
points(frontier_2002_ns$Volatility[three_pfs2002[1]], 
       frontier_2002_ns$Expected.Return[three_pfs2002[1]], 
       col="steelblue", pch=19)
points(frontier_2002_ns$Volatility[three_pfs2002[2]], 
       frontier_2002_ns$Expected.Return[three_pfs2002[2]], 
       col="steelblue", pch=19)
points(frontier_2002_ns$Volatility[three_pfs2002[3]], 
       frontier_2002_ns$Expected.Return[three_pfs2002[3]], 
       col="steelblue", pch=19)
par(mfrow=c(1,1))
```
Removing short selling again yields a very similar sight again as before.

We can clearly see that the marginal difference between 2001 and 2002 is minimal. 
A few differences with the outliers can be observed, the portfolio values stay
largely equal.
```{r}
par(mfrow=c(2,3))
barplot(frontier_2001_ns$Optimal.Weights[[three_pfs2001[1]]], 
        border=FALSE, col="black", main="1%")
barplot(frontier_2001_ns$Optimal.Weights[[three_pfs2001[2]]], 
        border=FALSE, col="black", main="1.5%")
barplot(frontier_2001_ns$Optimal.Weights[[three_pfs2001[3]]], 
        border=FALSE, col="black", main="2%")
barplot(frontier_2002_ns$Optimal.Weights[[three_pfs2002[1]]], 
        border=FALSE, col="black", main="1%")
barplot(frontier_2002_ns$Optimal.Weights[[three_pfs2002[2]]], 
        border=FALSE, col="black", main="1.5%")
barplot(frontier_2002_ns$Optimal.Weights[[three_pfs2002[3]]], 
        border=FALSE, col="black", main="2%")
par(mfrow=c(1,1))
```
Here we can clearly see how the weights change when not short-selling.
Individual stocks are bought much more to counteract the missing 
value in the short-sold stocks.
```{r}

weight_change_ns <- c(mean(abs(frontier_2001_ns$Optimal.Weights[[three_pfs2001[1]]]
                               -frontier_2002_ns$Optimal.Weights[[three_pfs2002[1]]])),
                   mean(abs(frontier_2001_ns$Optimal.Weights[[three_pfs2001[2]]]
                            -frontier_2002_ns$Optimal.Weights[[three_pfs2002[2]]])),
                   mean(abs(frontier_2001_ns$Optimal.Weights[[three_pfs2001[3]]]
                            -frontier_2002_ns$Optimal.Weights[[three_pfs2002[3]]])))


barplot(rbind(weight_change, weight_change_ns),beside = TRUE, 
        border=FALSE, col=c("steelblue","coral"), 
        legend.text = c("Short selling","No short selling"))
```

Short selling improves the expected returns of a portfolio considerably.

