---
title: "EDA - Home Assignment 1"
author: "Benjamin Meixner, 12302260"
date: "2025-03-19"
output: 
  pdf_document:
    latex_engine: tectonic
---

# Exercise 1

```{r}
data('mtcars')
head(mtcars)

mean(mtcars$mpg)
mean(mtcars$mpg)/3.78*1.609
```

Mean miles per gallon is 20.09, so about 8.55 kilometers per liter for the metric lovers in this course. :)

# Exercise 2

```{r}
mean(mtcars$am)
t.test(mtcars$mpg ~ mtcars$am, conf.level=0.95)
```

- null hypothesis: there is no difference between automatic transmission and fuel consumption
- alt hypothesis: there is difference between automatic transmission and fuel consumption

We reject the null, since p is very small and R also tells us that the alt hypothesis is to be used.

Also we can tell by the mean of `mtcars$am` that 40.6% of cars in this dataset have automatic transmission.

# Exercise 3

```{r}
ols <- lm(mtcars$mpg ~ mtcars$hp)
summary(ols)
```

From the coefficient of -0.06 we can tell that HP does not have an effect on fuel consumption.

```{r}
par(mfrow=c(2,2))
plot(ols)
```

From the Residuals vs Fitted plot I gather that a the actual data points are rather spread out and
the model does not fit as well. I can also tell that by the $R^2$ of 0.6 which is not so great.

I am only slightly confused by the plot in general.
