---
title: "EDA - Home Assignment 2"
author: "Benjamin Meixner, 12302260"
date: "2025-03-27"
output: 
  pdf_document:
    latex_engine: tectonic
---

# Linear Model
```{r}
library(ggplot2)
library(dplyr)

data <- read.csv("data/CEOdata.csv")
head(data)

sales_salary <- lm(salary ~ sales, data=data)

data %>%
    ggplot(aes(x = sales, y = salary)) +
    geom_point() +
    geom_smooth(method = "lm", se = F)

summary(sales_salary)
```

# Log Linear Model

```{r}
log_sales_salary <- lm(log(salary) ~ log(sales), data=data)
summary(log_sales_salary)

data %>%
    ggplot(aes(x = log(sales), y = log(salary))) +
    geom_point() +
    geom_smooth(method = "lm", se = F)

```

Here the points are closer together and the outliers are less extreme.
The points now are better-suited for a linear regression model.

# Graduate School

```{r}
grad_sales <- lm(salary ~ grad, data=data)
summary(grad_sales)
```

The slope of the graduate school LM model is negative. 
Grad School is not a good investment for CEOs.

# Combined Analysis

```{r}
combined <- lm(log(salary) ~ log(sales) + log(mktval) + profits + age + ceoten, data=data)
summary(combined)
```

The highest adjusted $R^2$ I could achieve was 0.30. Not a good model.
With logarithmic profits the $R^2$ score increased, but then the
result is quewed because it does not take profits into account when 
they are negative, since the log for a legative profit is not defined.

The best predictor for CEO compensation using my model is `sales`.

The problem with including `age` and `ceoten` in the model is that
the variables are influencing one another. Older people are 
more likely to spend more years in CEO positions.
