---
title: "EDA - Home Assignment 3"
author: "Benjamin Meixner, 12302260"
date: "2025-04-03"
output: 
  pdf_document:
    latex_engine: tectonic
---

```{r}
load("data/admissions.Rdata")
head(admissions)

lm1 <- lm(admit ~ gpa, data=admissions)
summary(lm1)

summary(admissions$gpa[admissions$admit == 1])
summary(admissions$gpa[admissions$admit == 0])
```

From the output of summary we can see that the correlation is not effective, 
since it only explains $R^2$ = 3.1\%$ of the data.

Also from the medians we can only see a $0.2$ point improvement of the students
admitted vs the non-admitted. I would not deem this hard evidence.

```{r}
admissions$gre = admissions$gre / 10
admissions$rank = factor(admissions$rank)
head(admissions)

glm1 <- glm(admit ~ gpa + gre + rank, data=admissions)
summary(glm1)
```

From the output of the Logit model I conclude that 
a bad ranking undergraduate school has a negative effect
while a high-ranking school does not seem to be a benefit per se.
Also here $20\%$ of the movement is explained by the model.
Still not great, but better than above.

```{r}
library("margins")
margins(lm1)
margins(glm1)
```

From the marginal effects of both models we can gather:

- when including just the GPA then a 1 point increase in GPA results in a $21.8\%$ higher acceptance rate.
- when including all variables then a 1 point increase in GPA results in a $15.5\%$ higher acceptance rate.

