---
title: "Empirical Data Analysis Homework 6"
author: "HH"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r loading data and required libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(estimatr)

# clean the working environmento
rm(list = ls())

# set your working directory
tryCatch({
  library(rstudioapi)
  setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
  }, error=function(cond){message(paste("cannot change working directory"))
})

# df <- read.csv([FILENAME])
df <- read.csv("data/guns.csv")
```

```{r Estimation, message=FALSE, warning=FALSE}
m1_robust <- lm_robust(log(vio) ~ shall, data = df, clusters = shall)
summary(m1_robust)
m2_robust <- lm_robust(log(vio) ~ shall + incarc_rate + density + avginc + pop + pb1064 + pw1064 + pm1029, data = df, clusters = shall)
summary(m2_robust)
```

We can see that the shall in (1) is negative, so that means we have a lower crime rate for states with shall laws. When comparing to the analysis in (2)
we can also see that the coefficient has decresed (-.0443 to -0.368) which means some of that variation is due to other factors than purely
the shall laws.

Also the significance level in (2) is not enough to reject the null, since the CI includes 0 and the p-value is 16%, too high.

```{r}
m3_robust <- lm_robust(log(vio) ~ shall + incarc_rate + density + avginc + 
                        pop + pb1064 + pw1064 + pm1029 + factor(year) + factor(stateid), 
                      data = df, clusters = shall)
summary(m3_robust)
```

Here I am not sure if I used the right constellation of yaer and stateid, but if I just focus on the `shall` variable:

```
Coefficients:
                    Estimate Std. Error   t value Pr(>|t|)   CI Lower  CI Upper
(Intercept)        3.972e+00  2.0693555   1.91944  0.27675 -14.811247 22.755242
shall             -2.799e-02  0.0673373  -0.41572  0.74223  -0.670145  0.614158
...omitted
```

I can clearly see that the effect has almost vanished. The p-value is 74% and the different years and also the different
states are much more of a concern than the shall law iteslf. I deem the effects of shall laws neglibile to the point one cannot
statistically prove they are working (so they probably are not).


