---
title: "l6_2"
author: "HH"
date: ''
output: html_document
editor_options: 
  chunk_output_type: console
---

# Difference in differences (Diff-in-diff, DID)
## in-class example: lecture 6
```{r}
# install.packages("estimatr")
library(ggplot2)
library(dplyr)
library(estimatr)  ### For lm_robust function (Clustered SE)

# Set your working directory

# load the csv file
df <- read.csv("MLDA 18-20y.csv")
```

# Focus on only two states
- subset data of state40 and state 41
- State --> Categorical variable
- Graph mortality rate in two states
```{r}
head(df)

df <- df %>% subset(state == 40 | state == 41)

# different ways of the line above
# df <- df[df$state == 41| df$state == 40, ]
# df <- df %>% filter(state == 41 | state == 40)

# making it a categorical variable
df$state <- as.factor(df$state)

ggplot(df, aes(x = year, y = mrate, color = state)) +
  geom_line()

```

# Running two by two DID (Two entities, pre/post treatment)
Generate a dummy indicating the time after the law change (post-treatment period).
The coefficient of the interaction term is the DID estimator.
```{r}

df$post <- ifelse(df$year >= 1975, T, F)

# Relevel the categorical variable --> we're going to use state 41 as the baseline. 
df$state <- relevel(df$state, ref = 2)
#df$state <- relevel(df$state, ref = "41")


m1 <- lm(mrate ~ state * post, data = df )
summary(m1)

# Why? serial correlation
### Robust model / clustered SE model 
m1_robust <- lm_robust(mrate ~ state * post, data = df, clusters = state)
summary(m1_robust)
```

# Another way to define the state40 dummy variable is the method we used in class.
```{r}
df$state40 <- ifelse(df$state == "40", TRUE, FALSE)

df$post <- ifelse(df$year > 1974, TRUE, FALSE) 

m1 <- lm(mrate ~ state40 *  post , data = df)
summary(m1)

# Why? serial correlation
### Robust model / clustered SE model 
m1_robust <- lm_robust(mrate ~ state40 * post, data = df, clusters = state)
summary(m1_robust)

```

# Reload the dataset to look at reforms changing the MLDA in more states

```{r}
# reload the data
df <- read.csv("MLDA 18-20y.csv")
# state --> categorical variable
df$state <- as.factor(df$state)

# base R histogram
hist(df$legal)

# histogram in ggplot
ggplot(df, aes(x=legal)) + 
  geom_histogram()

# histogram in ggplot by fequency
ggplot(df, aes(x=legal)) + 
  geom_histogram(aes(y = ..density..))

```

# run a simple regression 
mrate against legal
+ Run a regression accounting for the fact that there could be serial correlation between the observations within a state.
```{r}

m2 <- lm(mrate ~ legal, data = df)
summary(m2)

# a regression accounting for the fact that there could be serial correlation between the observations within a state.
m2_robust <- lm_robust(mrate ~ legal, data = df, clusters = state)
summary(m2_robust)
```

## State and time fixed effects
* Run a regression with clustered SEs which takes into account state and time fixed effects.
```{r}
## Note: time -> categorical variable

m3_robust <- lm_robust(mrate ~ state + as.factor(year) + legal, clusters = state, data = df)
summary(m3_robust)

```

# Common trend assumption 
Check visually if the common trends assumption is supported.
```{r}
ggplot(df, aes(x = year, y = mrate, color = state)) + 
  geom_line()


# Take a look at states 32 and 50 - there is too much noise in the full sample.
# State 32 has MLDA of 21, and state 50 has MLDA of 18 for the entire time
ggplot(df %>% subset(state == 50 | state == 32), aes(x = year, y = mrate, color = state)) + 
  geom_line()

```

# Adding state specific linear trends.
Run a regression with clustered SEs which takes into account state and time fixed effects and state-specific linear trends.
```{r}

# This is overkill for a sample of this size!
m4_robust <- lm_robust(mrate ~ state + as.factor(year) +  state:year + legal, data = df, clusters = state)
summary(m4_robust)

```