---
title: "EDA - Project"
author: "Ana Carolina Alencaster Herrero"
date: "2025-05-05"
output: pdf_document
---
 
Preparing the Dataset to be used more easily

```{r}
# load libraries
library(dplyr)
library(ggplot2)

rm(list = ls())

tryCatch({
  library(rstudioapi)
  setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
  }, error=function(cond){message(paste("cannot change working directory"))
})

df <- read.csv("project.csv")
df <- df[-(1:2), ] # first 2 lines are useless

# delete unfinished surveys (1)
table(df$Finished)
df <- df %>% filter(Finished != 0.0)
table(df$Finished)

# RENAME COLUMNS as the questions
df <- df %>% rename(journal_treatment = Q11,video_treatment = Q13, 
                    control_treatment = Q12, life_goals = Q1, 
                    set_goals = Q3_1, clear_sense = "Q2._1", 
                    write_goals = Q4, whatiwant = Q5_1, watch_content = Q6, 
                    age = Q7, gender = Q8)

table(df$gender)

# Create treatment variable based on which column has value 1
df$treatment <- case_when(
  df$journal_treatment != "" ~ "journal",
  df$video_treatment != "" ~ "video",
  df$control_treatment != "" ~ "control",
  TRUE ~ NA_character_  # Default case for any unmatched values
)
df$treatment <- factor(df$treatment, levels = c("control", "journal", "video"))

df$age <- as.numeric(df$age)
df$watch_content <- as.factor(df$watch_content)
df <- df %>% filter(gender != "")
df$gender <- as.factor(df$gender)
df$whatiwant <- as.numeric(df$whatiwant)
df$clear_sense <- as.numeric(df$clear_sense)
df$write_goals <- as.factor(df$write_goals)
df$set_goals <- as.factor(df$set_goals)
df$life_goals <- as.factor(df$life_goals)

# remove NAs in the dataset
df <- na.omit(df)
```

Begin Analysis

**DEMOGRAPHICS - we check if the randomization was done correctly**
**Overall:** we see the mean age is similar and the proportion of each gender is also similar - randomization was good.

```{r}

# AGE
#transform age into a numerical variable and compare means

#TODO run t.test
df %>%
  group_by(treatment) %>%
  summarise(mean_age = mean(age))

age_anova <- aov(age ~ treatment, data = df)
summary(age_anova) # insignificant

# Visualize age distribution by treatment
boxplot(age ~ treatment, data = df, 
        main = "Age Distribution by Treatment Group",
        xlab = "Treatment", ylab = "Age")

#GENDER
#Compare the count and proportion of people in each gender in the data
#count observations per group

gender_data <- df %>%
  group_by(treatment) %>%
  count(gender)

# Plot - COUNT
ggplot(gender_data, aes(x = treatment, y = n, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Gender Count by Treatment",
       x = "Treatment Group",
       y = "Count",
       fill = "Gender") +
  theme_minimal()

# slight female bias

#plotting proportion is better
gender_prop_data <- df %>%
  group_by(treatment, gender) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(treatment) %>%
  mutate(prop = n / sum(n))

ggplot(gender_prop_data, aes(x = treatment, y = prop, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Proportion of Genders by Treatment Group",
       x = "Treatment Group",
       y = "Proportion",
       fill = "Gender") +
  scale_y_continuous(labels = scales::percent_format()) +  # shows percent labels
  theme_minimal()
# female: 60%
# male: 40%
```

**Clear Sense**

Testing difference in means between groups

```{r}
# almost significant with 0.35
cs_anova <- aov(clear_sense ~ treatment, data = df)
summary(cs_anova)

# mean and median clarity
df %>%
  group_by(treatment) %>%
  summarise(mean_clarity = mean(clear_sense))
df %>%
  group_by(treatment) %>%
  summarise(median_clarity = median(clear_sense))

df %>% ggplot(aes(x = treatment, y = clear_sense, fill = treatment)) +
  geom_boxplot() +
  labs(title = "Clear Sense by Treatment",
       x = "Treatment Group",
       y = "Count",
       fill = "Treatment") +
  theme_minimal()
```

## Regression

**basic regression:** insignificant
**lognormal regression:** *still insignificant*

```{r}
#treatment as a dummy variable - we run some regressions
hist(df$clear_sense)

# basic
reg1 <- lm(clear_sense ~ treatment, df)
summary(reg1)

# logarithm
reg2 <- lm(log(clear_sense + 1) ~ treatment, df)
summary(reg2)

#scaled - also insignificant
reg3 <- lm(scale(df$clear_sense) ~ treatment, df)
summary(reg3)
```

**Adding Variables**
***regression with all variables***
-   [Overall:]{.underline}
    -   p-value: small
        ```         
         1.907e-09
        ```
    -   Adjusted R-Squared: 44.69
-   Shows -
    -   **Journal condition -** significant - negative
    -   **Video -** significant - negative
-   Other possible explanations
    -   what I want  - significant - positive
    -   set_goals - significant - positive
**Possible interpretation**
-   Maybe doing a journaling task or watching a video makes people more self-aware and therefore makes them think they have less clarity ab their life goals?
-   having set goals means you have more clarity about your life goals

```{r}
reg_allvariables<-lm(logclearsense ~ treatment + as.factor(watch_content) + age + as.factor(gender) + as.numeric(whatiwant) + as.factor(write_goals) + set_goals + as.factor(life_goals), df)
summary(reg_allvariables)
```

**Other types of regression**

```{r}

#could it be a polynomial regression? - how do we know if this is better?
reg_allvariables2<-lm(logclearsense ~ treatment^2 + as.factor(watch_content)^2 + age^2 + as.factor(gender)^2 + as.numeric(whatiwant)^2 + as.factor(write_goals)^2 + as.factor(set_goals)^2 + as.factor(life_goals)^2, df)
summary(reg_allvariables2)

```

**Adding interaction effects**

**EXPLORATORY FACTOR ANALYSIS**

**GAUSS-MARKOV ASSUMPTIONS**

```{r}
#GAUSS-MARKOV ASSUMPTIONS

#we have our basic model
model <- lm(clear_sense ~ treatment, data = df)
summary(model)

# 1. Linearity & Homoskedasticity
plot(model$fitted.values, model$residuals)  # Check for funnel shape 
abline(h = 0, col = "red")

library(lmtest)
bptest(model) #we see: p-value = 0.04241 - this means that we reject the null hypothesis that the data is homeoscedastic - WE MUST USE ROBUST STANDARD ERRORS?


# 2. Multicollinearity
library(car)
vif(reg_allvariables) #we see all adjusted values have an adjusted GVIF below 2 - so no concern for multicollinearity

#scale the data???

#rescaling the data: Z-score
Journal.sc<-EDA_Data_valuesnumeric[-1,]
Journal.sc<-Journal.sc[-5,]
Journal.sc<-Journal.sc[,-22] #quito clear_sense?
Journal.sc<-Journal.sc[,21:27]

Journal.sc <- Journal.sc %>% rename(life_goals = Q1, 
                                    set_goals = Q3_1, 
                                    write_goals = Q4, whatiwant = Q5_1, watch_content = Q6, 
                                    age = Q7, gender = Q8)


Journal.scaled<-scale(Journal.sc) #we select the values for each category and calculate their z-score using scale(), and then we index these scaled values and define them as a new variable
summary(Journal.sc)
cor_matrix <- cor(Journal.sc)

# View the matrix
library(psych)
corr.Journal.sc <- round(cor(Journal.scaled),3) #we index the calculated zscore values and calculate the correlation, then round to 3 digits
corPlot(corr.Journal.sc,
        numbers=T,diag=F,upper=F,
        main="Correlations between variables" ) 

KMO(Journal.scaled) #overall MSA is >0.5 - this indicated that there is some correlation between the variables

#EIGENVALUES
#how many factors should we actually have  - the first 3 are >1 (life_goals, set_goals, write_goals)

library(nFactors) #CFA
eigen(corr.Journal.sc)
#first 3 are >1 - but CAREFUL - one is clear_sense

Journal.sc

journal_efa<- factanal(x=Journal.sc, factors=3, rotation="varimax", scores="Bartlett")


#Running the Regression With Factors

factors<-journal_efa$scores


#if we wanted to relate eth into factors

#factor 1 - age
#factor 2 - write_goals, watch_content  
#factor 3 - set_goals, whatiwant


factors <- journal_efa$scores

colnames(factors) <- c(
  "age_factor",
  "write_goals_watch_content",
  "set_goals_whatiwant"
)

```

**Results from the EFA Regression**
```{r}
Journal_reggression_data <- data.frame(clear_sense2 = df$clear_sense, factors)

model5 <- lm(clear_sense2 ~ age_factor + write_goals_watch_content + set_goals_whatiwant, data = Journal_reggression_data)
summary(model5)

#what if now we add the treatments to this model?

Journal_reggression_data_complete <- data.frame(clear_sense2 = df$clear_sense, factors, treatment = df$treatment)

model6 <- lm(clear_sense2 ~ age_factor + write_goals_watch_content + set_goals_whatiwant + treatment, data = Journal_reggression_data_complete)
summary(model6)


head(efa_v$scores)
reg4<-lm(age ~ as.factor(watch_content), df)
summary(reg4) #ppl who are older never watch content?????????

# Create the barplot ????????????
ggplot(df, aes(x = treatment, y = age, fill = treatment)) +
  geom_bar(stat = "identity") +
  labs(title = "Age by treatment", x = "Group", y = "Age")

plot(log(df$clear_sense))
```

```{r}
#WATCH CONTENT
table(df$watch_content)

watchcontent_data <- df %>%
  group_by(treatment) %>%
  count(watch_content)

ggplot(watchcontent_data, aes(x = treatment, y = n, fill = watch_content)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Content Watching Habits by Treatment Group",
       x = "Treatment Group",
       y = "Count",
       fill = "Watch Content")
```

```{r}

```
