---
title: "Quantitative Methods 2 - Case Study 3"
author: "Group Number: [Your Group Number Here]"
date: "May 31, 2024"
output: pdf_document
---

```{r setup, include=FALSE}
# Loading necessary packages
library(tidyverse)
```

## 1. Computing the Log-Return Series

First, we load the historical price series data for IBM and compute the log-return series.

```{r}
# Load the data
ibm_data <- read.csv("IBMdata.csv")

# Compute the log-return series
log_returns <- diff(log(ibm_data$Price))

# Plot the histogram of log-returns
hist(log_returns, main="Histogram of Log-Returns", xlab="Log-Returns", breaks=50, col="blue")
```

## 2. Calculating Sample Statistics

Next, we calculate the sample mean (\(\bar{x}\)) and sample standard deviation (\(s\)) of the log-returns. These statistics will serve as estimators for the unknown parameters \(\mu\) and \(\sigma\).

```{r}
# Calculate sample mean and standard deviation
mean_log_return <- mean(log_returns)
sd_log_return <- sd(log_returns)

# Report the results
mean_log_return
sd_log_return
```

## 3. Generating Realizations of the Random Variable \(Z\)

We set \(n = 100000\) and use the function `rnorm` to generate \(n\) realizations of the random variable \(Z\).

```{r}
# Parameters
n <- 100000
S0 <- 110

# Generate n realizations of Z
Z <- rnorm(n)

# Calculate SF realizations
SF <- S0 * exp((mean_log_return - (sd_log_return^2) / 2) * 1 + sd_log_return * sqrt(1) * Z)
```

## 4. Calculating the Payoff Realizations

We set \(K1 = 100\) and \(K2 = 120\), and use the payoff function to obtain realizations for the payoff.

```{r}
K1 <- 100
K2 <- 120

# Payoff function
payoff <- SF + pmax(K1 - SF, 0) - pmax(SF - K2, 0)

# Plot the histogram of payoffs
hist(payoff, main="Histogram of Payoffs", xlab="Payoffs", breaks=50, col="green")
```

## 5. Approximating the Value \(V\) Using the Monte Carlo Method

We use the generated realizations to approximate the value \(V\).

```{r}
# Approximate the value V
V_hat <- mean(payoff)
V_hat
```

## 6. Sample Variance of \(V\)

We calculate the sample variance of \(V\).

```{r}
# Sample variance
var_payoff <- var(payoff)
var_payoff
```

## 7. Analysis of Variance and Sample Size

We discuss how the variance of \(V\) changes as we increase the sample size \(n\).

```{r}
# The variance of the estimate V_hat
var_V_hat <- var_payoff / n
var_V_hat

# Discussion:
# As the sample size n increases, the variance of V_hat decreases.
# This is in line with the Central Limit Theorem, which states that the distribution of the sample mean will
# tend to be normal with a variance equal to the population variance divided by the sample size.
```

## 8. Calculating Profit/Loss (P/L) Values and Value at Risk (VaR)

We calculate the P/L values and estimate the 5% quantile (VaR).

```{r}
# Initial cost
C <- 110

# Calculate P/L values
PL <- payoff - C

# Plot the histogram of P/Ls
hist(PL, main="Histogram of Profit/Loss", xlab="Profit/Loss", breaks=50, col="red")

# Estimate expected P/L
expected_PL <- mean(PL)
expected_PL

# Calculate the 5% quantile (VaR)
VaR_5 <- quantile(PL, 0.05)
VaR_5

# Discussion on VaR:
# The 5% quantile (VaR) represents the maximum loss that can be expected with 95% confidence.
# In other words, there is a 5% chance that the loss will exceed this value.
```

