---
title: "Quantitative Methods 2 - Case Study 2"
author: "Group Number: [Your Group Number Here]"
date: "June 7, 2024"
output: pdf_document
---

```{r setup, include=FALSE}
# Loading necessary packages
library(tidyverse)
library(ggplot2)
library(dplyr)
```

## 1. Descriptive Analysis of the Data Set

First, we load the dataset and provide a descriptive analysis of the variables.

```{r}
# Load the data
wines <- read.csv("wines.csv", sep = "")

# Basic information about the dataset
str(wines)
summary(wines)

# Sample size
sample_size <- nrow(wines)
sample_size

# Plot the distribution of wine scores
ggplot(wines, aes(x = points)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Wine Scores", x = "Score", y = "Count")

# Plot the distribution of wine prices
ggplot(wines, aes(x = price)) +
  geom_histogram(binwidth = 5, fill = "green", color = "black") +
  labs(title = "Distribution of Wine Prices", x = "Price (USD)", y = "Count")

# Plot the distribution of countries of origin
ggplot(wines, aes(x = country)) +
  geom_bar() +
  xlab("Country") +
  ylab("Quantity") +
  ggtitle("Number of Wines from each Country")

# Plot the distribution of provinces of origin
ggplot(wines, aes(x = province)) +
  geom_bar() +
  xlab("Province") +
  ylab("Quantity") +
  ggtitle("Number of Wines from each region") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

# Plot the distribution of wines' variety
ggplot(wines, aes(x = variety)) +
  geom_bar() +
  xlab("Variety") +
  ylab("Quantity") +
  ggtitle("Number of Wines of each variety") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

# Plot the distribution of wineries of origin
# Count the number of wines per winery
winery_counts <- wines %>%
  group_by(winery) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  top_n(20, count)

# Create the histogram
ggplot(winery_counts, aes(x = reorder(winery, -count), y = count)) +
  geom_bar(stat = "identity") +
  xlab("Winery") +
  ylab("Quantity") +
  ggtitle("Number of Wines from Top 20 Wineries") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

```

## 2. Analysis of Sample on Country and Variety Level

### 2(a). Comparing Countries

We compare the average prices and average quality scores of wines from France and Italy.

```{r}
# Average price by country
avg_price_by_country <- wines %>%
  group_by(country) %>%
  summarize(mean_price = mean(price, na.rm = TRUE))

avg_price_by_country

# Average quality by country
avg_quality_by_country <- wines %>%
  group_by(country) %>%
  summarize(mean_quality = mean(points, na.rm = TRUE))

avg_quality_by_country
```

### 2(b). Comparing Wine Varieties

We determine which variety of wine is the most expensive and which has the highest average quality.

```{r}
# Average price by variety
avg_price_by_variety <- wines %>%
  group_by(variety) %>%
  summarize(mean_price = mean(price, na.rm = TRUE)) %>%
  arrange(desc(mean_price))

avg_price_by_variety

# Average quality by variety
avg_quality_by_variety <- wines %>%
  group_by(variety) %>%
  summarize(mean_quality = mean(points, na.rm = TRUE)) %>%
  arrange(desc(mean_quality))

avg_quality_by_variety
```

### 2(c). Combination of Price and Points

We analyze the combinations of price and points to find the highest average price and best average quality.

```{r}
# Highest average price and best average quality by country and variety
avg_price_quality <- wines %>%
  group_by(country, variety) %>%
  summarize(mean_price = mean(price, na.rm = TRUE),
            mean_quality = mean(points, na.rm = TRUE)) %>%
  arrange(desc(mean_price), desc(mean_quality))

avg_price_quality
```

## 3. Testing the Sommelier's Claims

We test the claims about French Chardonnay.

### 3(a). Proportion Test for French Chardonnay in Champagne

```{r}
# Subset French Chardonnays
french_chardonnay <- subset(wines, country == "France" & variety == "Chardonnay")

# Proportion of French Chardonnay in Champagne
prop_champagne <- mean(french_chardonnay$province == "Champagne")
prop_champagne

# Hypothesis test for proportion
prop.test(x = sum(french_chardonnay$province == "Champagne"), n = nrow(french_chardonnay), p = 0.11)
```

### 3(b). Mean Price Test for French Chardonnay

```{r}
# Mean price of French Chardonnay
mean_price_chardonnay <- mean(french_chardonnay$price, na.rm = TRUE)
mean_price_chardonnay

# t-test for mean price
t.test(french_chardonnay$price, mu = 55.5)
```

## 4. Comparison of Price and Quality Between Countries

### 4(a). Boxplot of Wine Prices by Country

```{r}
# Boxplot of prices by country
ggplot(wines, aes(x = country, y = price)) +
  geom_boxplot(fill = c("blue", "green")) +
  labs(title = "Boxplot of Wine Prices by Country", x = "Country", y = "Price (USD)")
```

### 4(b). Boxplot of Wine Quality by Country

```{r}
# Boxplot of quality by country
ggplot(wines, aes(x = country, y = points)) +
  geom_boxplot(fill = c("blue", "green")) +
  labs(title = "Boxplot of Wine Quality by Country", x = "Country", y = "Quality (Points)")
```

### 4(c). Confidence Intervals for Wine Prices

```{r}
# Confidence intervals for wine prices
ci_price_france <- t.test(wines$price[wines$country == "France"])$conf.int
ci_price_italy <- t.test(wines$price[wines$country == "Italy"])$conf.int

ci_price_france
ci_price_italy

# Interpretation:
# If the confidence intervals do not overlap, there is statistical evidence of a difference in wine prices.
```

### 4(d). Confidence Intervals for Wine Quality

```{r}
# Confidence intervals for wine quality
ci_quality_france <- t.test(wines$points[wines$country == "France"])$conf.int
ci_quality_italy <- t.test(wines$points[wines$country == "Italy"])$conf.int

ci_quality_france
ci_quality_italy

# Interpretation:
# If the confidence intervals do not overlap, there is statistical evidence of a difference in wine quality.
```

### 4(e). Combined Interpretation

```{r}
# Combined interpretation of price and quality confidence intervals
# Discussion on whether the results are expected or surprising based on confidence intervals for price and quality.
# Also considering the Central Limit Theorem and assumptions made during tests.
```

