---
output:
  pdf_document:
    latex_engine: tectonic
    fig_caption: yes
---

\thispagestyle{empty}

\begin{center}
  \vspace*{2cm}
  \Huge\textbf{Case Study 4} \\[0.5cm]
  \Large\textbf{Linear Regression of CEO Compensations on the Firm’s Sales} \\[1cm]
  \large
  \begin{tabular}{c}
    Benjamin Maixner \\
    Chirill Glitos \\
    Franziska Kirschner \\
    Said Rizvanov \\
  \end{tabular}
  \vfill
  \normalsize
  \textbf{Date}: `r format(Sys.Date(), "%B %d, %Y")`
\end{center}

\newpage

## Setting up

First we need to set up our environment by loading relevant libraries and importing the 
data we need from the given CSV.

Then we also draw histograms to first visualize the data and get a feel for the 
dataset we are dealing with.

```{r setup, include=TRUE}
# Load necessary libraries
library(ggplot2)
library(readr)

# Extracting data from the .csv file
data <- read.csv("CEO_compensations.csv", sep = ",")

# Attach the dataset to use variable names directly
attach(data)

# Histograms for X and Y
par(mfrow=c(1,2))  #plotting layout
hist(SALES, main="Histogram of Sales", xlab="Sales", col="blue", breaks=50)
hist(COMP, main="Histogram of Compensation", xlab="Compensation", col="green", breaks=50)
```

The histogarms seem to be aligning quite well with one another. In the lower part with more data points as
well as in the higher regions with considerably less density of data points.

To get a better overview we also plot a logarithmic scale histogram, since this will be used later for a more 
accurate regression approximation.

```{r}
# Log-transformed histograms
par(mfrow=c(1,2))  #plotting layout
hist(log2(SALES), main="Histogram for Sales", xlab="log2(Sales)", col="blue", breaks=50)
hist(log2(COMP), main="Histogram for Compensation", xlab="log2(Compensation)", col="green", breaks=50)
```

## Regression

To start solving the task we first implement a standard linear regression model.

```{r}
# Simple linear regression without log-transform
model1 <- lm(COMP ~ SALES)
summary(model1)
```

The summary tells us that our approach is already quite good, since we see 3 stars `***` with both
variables. This signals high significance and therefore a close correlation between the 2 variables
*Sales* and *Income*.

But we think we can do better by using a logarithmic scale to reduce the spread of the data.

```{r}
# Simple linear regression with log-transform
model2 <- lm(log2(COMP) ~ log2(SALES))
summary(model2)
```

Looking the summary now we see an even higher significance, since we don't even see any zeroes 
with the `SALES` variable, but also an exponential notation, meaning the value is even smaller
and thus resulting in a better approximation model.

## Scattering the Plot

Now to get an even better feel of our approximation we will use scatter plots to show the data set 
as well as the regression line.

```{r}
# Scatter plots with regression lines
par(mfrow=c(1,2))  # 1x2 plotting layout

# Scatter plot for original data
plot(SALES, COMP, main="Compensation vs. Sales", xlab="Sales", ylab="Compensation")
abline(model1, col="red")

# Scatter plot for log-transformed data
plot(log2(SALES), log2(COMP), main="log2(Compensation) vs. log2(Sales)", xlab="log2(Sales)", ylab="log2(Compensation)")
abline(model2, col="red")
```

Comparing the 2 plots we see a steeper red linear regression line. Also the spread of the data points
is much smoother using the logarithm. More spread out in the clustering region, but also closer together in the outer
regions.

```{r}
# R-squared for the models
r_squared_model1 <- summary(model1)$r.squared
r_squared_model2 <- summary(model2)$r.squared

r_squared_model1
r_squared_model2
```

And we can also verify that the spread has decreased as is indicated by the `Multiple R-squared` in the summary. 
It has dramatically improved from `0.141` to `0.2416`. Almost double, therefore half the spread overall.


```{r}
# Choosing the final model
chosen_model <- model2  # Assuming log-transformed model was chosen

# Interpretation of the slope
slope <- coef(chosen_model)[2]
slope
```

For the reason of less spread and a more accurate approximation we chose the logarithmic model 
to base our predictions on.

## Tell me Crystal Sphere

Lastly we want to finally reap the fruits of our effort and predict the compensation
of CEOs which are not included in the dataset.

```{r}
# Predicting compensation for sales of 1.6384 billion USD
sales_values <- seq(from=min(SALES), to=max(SALES), length.out=100)
predicted_log2_compensation <- predict(chosen_model, newdata = data.frame(log2.SALES. = log2(sales_values)))
predicted_compensation <- 2^predicted_log2_compensation
predicted_compensation
```

```{r}
# Perform statistical test on the chosen model
summary(chosen_model)
```


```{r}
# Standardize the variables
log2_sales_std <- scale(log2(SALES))
log2_compensation_std <- scale(log2(COMP))

# Fit the model with standardized variables
model_std <- lm(log2_compensation_std ~ log2_sales_std)
summary(model_std)
```


```{r}
# Verify that one of the coefficients is the correlation coefficient
correlation <- cor(log2_sales_std, log2_compensation_std)
coef_model_std <- coef(model_std)[2]

list(correlation = correlation, coef_model_std = coef_model_std)
```

