---
title: "BA2 Mock Exam"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction
In this exam, you will process a new dataset, perform exploratory data checks, conduct regression and PCA analyses, evaluate a portfolio, and calculate risk measures. Fill in the code chunks with your solutions.

---

## Exercise 1: Data Screening
- Load the provided dataset. Inspect it for missing values, logical errors, or outliers.
- Subset the data as needed (e.g., time periods or targeted columns).

```{r data-screening}
# Load libraries
# ...
# Read in data
# ...
# Summary statistics
# ...
# Data cleaning steps
# ...
```

---

## Exercise 2: Exploratory Visualization & Returns
- Plot key variables from the cleaned dataset.
- If applicable, compute daily or monthly returns. Check for data completeness.

```{r exploratory-visuals}
# Plot key variables (e.g., price, yields, etc.)
# ...
```

```{r returns}
# Convert price data to returns
# ...
# Summaries of returns
# ...
```

---

## Exercise 3: PCA on Yield Curve (or Other Multivariate Data)
- Standardize the relevant variables.
- Compute and interpret the principal components.
- Plot the loadings and the explained variance.

```{r pca}
# Perform PCA on standardized data
# ...
# Visualize loadings
# ...
# Visualize explained variance
# ...
```

---

## Exercise 4: CAPM / Factor Model Analysis
- Pick a subset of assets or stocks.
- Estimate betas and alphas over rolling windows.
- Plot the time series of betas (and alphas).

```{r capm}
# Merge dataset with market and risk-free data
# ...
# Rolling regression for CAPM
# ...
# Plot beta and alpha
# ...
```

---

## Exercise 5: Portfolio Analysis & Risk Measures
- Construct a portfolio of several assets.
- Calculate portfolio returns, weights, VaR, and Expected Shortfall.
- Plot results and interpret.

```{r portfolio}
# Define portfolio (assets, weights, etc.)
# ...
# Compute portfolio returns
# ...
# Compute portfolio VaR, ES
# ...
# Visualize results
# ...
```

---
