---
title: "Business Analytics II - Group 1 - Home Assignment 6"
author: "Yurii Masiuk, Vanda Lakatos, Kornel Auer, Vivien Arns, Benjamin Meixner, Nicholas Olsa"
date: "2025-01-03"
output: 
  pdf_document:
    latex_engine: tectonic
---
# Introduction

All members are able to solve exercises 1, 2 and 3 with all sub-points.

# EXERCISE 1
```{r, message=FALSE, warning=FALSE, error=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))
library(viridis)
library(xts)
library(lubridate)

yields <- read.csv("../data/data_yields.csv")
head(yields)
yields <- yields[ , c("Date", "X3M", "X1Y", "X5Y", "X7Y", "X10Y")]
head(yields)
```
## 1) PLOT YIELD CURVE FOR 2021-08-23
```{r}
maturities <- c(3/12, 1, 5, 7, 10)
plot(x=maturities, y=yields[1,-1],
     col="steelblue", type="o", pch=16, lty=2,
     xlab="Maturity (in years)", ylab="Yield p.a. (%)",
     main=paste0("German yield curve on ", yields[1,1]))
```
## 2) PLOT EVOLUTION OF YIELDS ACROSS MATURITIES
```{r}
# convert to xts and plot
yields.xts <- xts(yields[,-1], order.by=ymd(yields$Date))

yield_plot <- plot(yields.xts, col=viridis(n=ncol(yields.xts)),
                   main="Euro benchmark yields")
yield_plot <- addLegend("bottomleft",
                        legend.names=c("3M", paste0(c(1,5,7,10),"Y")),
                        fill=viridis(n=ncol(yields.xts)))
yield_plot
```
Generally speaking, the yields of different maturities exhibit a certain degree of co-movement over time, which is highly influenced by the general economic stability. In example, during the 2008 financial crisis, we can observe a significant drop in the short-term yields, while the long-term yields remained more stable. In the time period from 2004 to 2019 yields decreased gradually, even to negative levels.

## 3) CALCULATE PRINCIPAL COMPONENTS
```{r}
#1. drop first column containing dates
#2. standardize each column
yields_std <- apply(yields[,-1], 2, function(x) (x - mean(x))/sd(x))

# check mean of each standardized variable
# -> basically zero (just numerical imprecision)
colMeans(yields_std) 

# check standard deviation of each standardized variable
# -> exactly 1 for all variables
apply(yields_std, 2, sd)

# could omit last two arguments since we previously already scaled data by hand
pca <- prcomp(yields_std, center=TRUE, scale.=TRUE)

plot(pca$sdev^2/sum(pca$sdev^2),
     col="steelblue", type="b", pch=20,
     main="Proportion of explained Variance", ylab="% explained"
     ,xlab="Number of PCs")
abline(h=0, lty=2, col="darkgrey")
```
Based on the visualization, we can see that the first principal component explains the majority of the variance in the data. The second and third principal components explain a smaller portion of the variance. The fourth and fifth principal components explain a very small portion of the variance in the data. Because we try to pick the smallest number of PCs, we can either choose two or three PCs, which explain a majority of the variations in the model. The rest of the PCs explain a very small portion of the variance in the data and can be omitted.

## 4) APPROXIMATE YIELD CURVE WITH FIRST 3 PCs
```{r}
# get mean and SD of original data:
yields_mean <- colMeans(yields[,-1]) #drop 1st col with dates
yields_sd <- apply(yields[,-1], 2, sd)

# D* = X* Lambda'*
#note that we need to transpose Lambda (pca$rotation)!
yields_approx_3pc <- pca$x[,1:3] %*% t(pca$rotation)[1:3,]
head(yields_approx_3pc)

# scale to original SD and add back the mean
for(i in 1:ncol(yields_approx_3pc)){
  yields_approx_3pc[,i] <- yields_approx_3pc[,i] * yields_sd[i] + yields_mean[i]
}

# colnames(yields_approx_3pc) <- colnames(yields[,-1])
yields_approx_3pc.xts <- xts(yields_approx_3pc, order.by=rev(time(yields.xts)) )
# why rev() in the line above?
# -> yields (and therefore yieds_approx_3pc) are ordered from last to first,
# but yields.xts (and therefore time(yields.xts)) are automatically ordered from first to last
```
## 5) CHECK APPROXIMATION AGAINST ORIGINAL DATA
```{r}
plot_3pc <- plot(yields.xts[,c("X3M", "X5Y", "X10Y")],
                 col="magenta", multi.panel=TRUE, main="Approximation with 3 PCs")
plot_3pc <- lines(yields_approx_3pc.xts[,"X3M"],
                  col="aquamarine", on=1)
plot_3pc <- lines(yields_approx_3pc.xts[,"X5Y"],
                  col="aquamarine", on=2)
plot_3pc <- lines(yields_approx_3pc.xts[,"X10Y"],
                  col="aquamarine", on=3)
plot_3pc <- addLegend("bottomleft", legend.names=c("original", "approximation"),
                      col=c("magenta", "aquamarine"), lwd=1, on=3)
plot_3pc
```
The visualization proofs that the variations in yields can be accurately described using three PCs, we can draw this conclusion because the lines of the original and approximated data nearly coincide over the whole time frame.

# Exercise 2
## 1) VISUALLY SHOW LOADINGS FROM LOADINGS MATRIX LAMBDA FOR FIRST 3 PCs (in one plot)
```{r}
plot(x=maturities, y=pca$rotation[,1], col="steelblue", type="o",
     ylim=c(-1,1), ylab="loading", xlab="maturity (in years)")
lines(x=maturities, y=pca$rotation[,2], col="coral", type="o")
lines(x=maturities, y=pca$rotation[,3], col="forestgreen", type="o")
abline(h=0, lty=2, col="grey")
legend("bottomleft", legend=colnames(pca$rotation)[1:3], bty="n",
       lty=1, pch=1, col=c("steelblue", "coral", "forestgreen"))
```
The interpretation given to the first three PCs during the lecture also appear plausible for our data set. We observe that PC1 is uniformly related to yields of all maturities, it can be interpreted as the level of the yield curve.
For PC2, we see that loadings in column two flip the sign from the short to the long end of the yield curve. This suggests that PC2 captures the slope of the yield curve.
The third principal component, PC3, is related to the curvature of the yield curve. The loadings are negative for the short (3-month) and long end (10-years) of the yield curve, while they are positive for the middle maturities.

## 2) COMPARISON OF PCs AGAINST DIRECT MEASURES OF LEVEL, SLOPE, AND CURVATURE
```{r}
### 2a
# Calculate measure level as the difference between 5Y yield and the mean of all yields
measure_level <- (yields.xts[,"X5Y"]-yields_mean["X5Y"])/yields_sd["X5Y"]

# Plot the measure_level in blue
plot(measure_level, col="steelblue", type="l", 
     main="Comparison of Measure Level and 1st PC", 
     ylab="Standardized Value", xlab="Date")

# Overlay the standardized 1st PC in coral
lines(xts(pca$x[,1], order.by=ymd(yields$Date))/sd(pca$x[,1]), col="coral")

# Add a legend to differentiate the two series
legend("topright", legend=c("Measure Level", "1st PC"), 
       col=c("steelblue", "coral"), lty=1, bty="n")
```

This plot visualizes that the PC1 tracks the average level of the yields fairly well, which aligns with the interpretation of PC1 as the level of the yield curve.

```{r}
### 2b

# Calculate slope measure as the difference between 10Y and 3M yields
measure_slope <- yields.xts[,"X10Y"] - yields.xts[,"X3M"]

# Standardize the slope measure (center to mean 0, scale to standard deviation 1)
measure_slope <- (measure_slope - mean(measure_slope)) / sd(measure_slope)

# Plot the standardized slope measure in blue
plot(measure_slope, col="steelblue", type="l", 
     main="Comparison of Slope Measure and 2nd PC", 
     ylab="Standardized Value", xlab="Date")

# Overlay the standardized 2nd principal component (sign-adjusted)
lines(xts(pca$x[,2], order.by=ymd(yields$Date)) / sd(pca$x[,2]) * (-1), col="coral")

# Add a legend to distinguish between the two lines
legend("topright", legend=c("Measure Slope", "2nd PC (standardized)"), 
       col=c("steelblue", "coral"), lty=1, bty="n")

```

The plot shows that the second principal component tracks the slope of the yield curve. The negative sign of the second principal component is due to the fact that the loadings of the second principal component are negative for the short end of the yield curve and positive for the long end of the yield curve, which is visible in the differences of the lines and the change in sign across the time-frames. This is consistent with the interpretation of the second principal component as the slope of the yield curve like obtained in the visualization of exercise 2.1 (considering the sign-adjustment).

```{r}
### 2c
measure_curvature <- yields.xts[,"X10Y"] + yields.xts[,"X3M"] - 2*yields.xts[,"X5Y"]
measure_curvature <- (measure_curvature-mean(measure_curvature) ) / sd(measure_curvature)
plot(measure_curvature , col="steelblue")
lines(xts(pca$x[,3], order.by=ymd(yields$Date))/sd(pca$x[,3])*(-1), col="coral")
```
```{r}
# Calculate curvature measure: combines 10Y, 3M, and 5Y yields
measure_curvature <- yields.xts[,"X10Y"] + yields.xts[,"X3M"] - 2 * yields.xts[,"X5Y"]

# Standardize the curvature measure (center to mean 0, scale to standard deviation 1)
measure_curvature <- (measure_curvature - mean(measure_curvature)) / sd(measure_curvature)

# Plot the standardized curvature measure in blue
plot(measure_curvature, col="steelblue", type="l",
     main="Comparison of Curvature Measure and 3rd PC",
     ylab="Standardized Value", xlab="Date")

# Overlay the standardized 3rd principal component (sign-adjusted)
lines(xts(pca$x[,3], order.by=ymd(yields$Date)) / sd(pca$x[,3]) * (-1), col="coral")

# Add a legend to differentiate between the two lines
legend("topright", legend=c("Measure Curvature", "3rd PC (standardized)"),
       col=c("steelblue", "coral"), lty=1, bty="n")

```


# Exercise 3
## 1) PRESENT VALUE OF BOND PORTFOLIO AS OF 2021-08-23
```{r}
yields.xts["2021-08-23"]
maturities

cf <- c(5,5,15,8,25) #net cash flows
pv <- sum(cf/(1+as.vector(yields.xts["2021-08-23"])/100)^maturities)
pv
```
## 2) EXTREME YIELD CURVE STEEPENING
```{r}
#[a] convert 2nd PC to xts and take differences
pc2 <- xts(pca$x[,"PC2"], order.by=ymd(yields$Date))
d_pc2 <- diff.xts(pc2)

#[b]
pc2_var95 <- quantile(na.omit(d_pc2), probs=0.05) #5%, since we found positive loadings for short end and negative loadings for long end of the yield curve
```
## 3) TRANSLATE EXTREME CHANGE IN PC2 TO YIELD CHANGES ACROSS MATURIES
```{r}
yields_sd #original sd of yields
d_yields_var95 <- pc2_var95*t(pca$rotation)[2,]*yields_sd #change in PC2 translated to change in standardized yield data; then scaled back to changes on original scales
```
## 4) RE-EVALUATE THE PORTFOLIO WITH NEW YIELD LEVELS
```{r}
yields_var95 <- yields.xts["2021-08-23"] + d_yields_var95
pv_var95 <- sum(cf/(1+yields_var95/100)^maturities)
pv_var95
```
## 5) CHANGE IN PORTFOLIO VALUE
```{r}
pv_var95 - pv
```
